The effect of prenatal environment on brain metabolic function and perception in a lizard

### 2025-10-12 ms.qmd
Update ms.qmd

## Abstract
Cognitive processes such as the ability to perceive prey, are crucial for survival and reproduction. Early-life conditions like stress-related hormones or thermal environments, can shape cognitive abilities by influencing brain structure and function. Mitochondrial physiology is thought to be a central mechanism underlying these effects given that it is crucial for the energy production needed for brain function. Here, we investigated the combined influence of prenatal corticosterone (CORT)–a key stress hormone in reptiles–and incubation temperature on brain mitochondrial function and impacts on prey detection in the delicate skink (_Lampropholis delicata_). We manipulated egg CORT levels and incubation temperature, then assessed the ability of lizards to detect chemical and visual prey stimuli. Using flow cytometry, we measured metabolic function, reactive oxygen species (ROS) production, and oxidative stress in the olfactory bulbs and optic tecta-two brain regions involved in sensory processing. We predicted that CORT exposure and high temperatures would interact to shape perception of prey via changes in mitochondrial respiration and oxidative damage. While metabolic function remained robust to early life conditions, CORT and temperature interacted to influence oxidative stress in the optic tecta. Additionally, CORT-treated lizards responded faster when presented with chemical cues but not visual stimuli. Our findings suggest that prenatal conditions can have lasting effects on brain physiology and cognition, though the role of mitochondrial function in these processes remains unclear. Future research should explore alternative mechanisms that could link early-life stress to cell metabolism and cognitive performance.

>>>>

## Abstract
Cognitive processes such as the ability to perceive prey, are crucial for survival and reproduction. Early-life conditions like stress-related hormones or thermal environments, can shape cognitive abilities by influencing brain structure and function. Mitochondrial physiology is thought to be a central mechanism underlying these effects given that it is crucial for the energy production needed for brain function. Here, we investigated the combined influence of prenatal corticosterone (CORT)–a key stress hormone in reptiles–and incubation temperature on brain mitochondrial function and impacts on prey detection in the delicate skink (_Lampropholis delicata_). We manipulated egg CORT levels and incubation temperature, then assessed the ability of lizards to detect chemical and visual prey stimuli. Using flow cytometry, we measured metabolic function, reactive oxygen species (ROS) production, and oxidative stress in the olfactory bulbs and optic tecta-two brain regions involved in sensory processing. While metabolic function remained robust to early life conditions, CORT and temperature interacted to influence oxidative stress. Additionally, CORT-treated lizards responded faster when presented with chemical cues but not visual stimuli. Our findings suggest that prenatal conditions can have lasting effects on brain physiology and cognition, highlighting a surprising decoupling between mitochondrial function and perception in lizards.
